---
title: "Bacterial-fungal network calculation"
output:
  html_document:
    theme: united
    highlight: breezedark
    toc: yes
    toc_float: true
    df_print: paged
---

# Outline

- Calculate the network with all taxa for each SLR rank r (number of latent factors) and compute the EBIC.

- Export the StARS-selected edges with partial correlations as edge weights (Supplementary Table 11).

- Robustness: estimate networks on five random subsamples of 80% of the samples at the lambda selected for the full data, and calculate the relative Hamming distance (FP + FN) / P to the full-data network (Supplementary Table 14).

- The rank with the lowest EBIC among robust models (mean relative Hamming distance < 0.5) is used.

```{r eval=FALSE}
HOME_LOCAL <- "./"   # folder with the input files; output is written to HOME_LOCAL/SPIEC-EASI_robustness/
source(paste(HOME_LOCAL, "00_network_functions.R", sep = ""))
dir.create(paste(HOME_LOCAL, "SPIEC-EASI_robustness", sep = ""), showWarnings = FALSE)
```

# Calculate networks

### Specify input

```{r eval=FALSE}
library(phyloseq)

## specify input files
physeq1 <- readRDS(paste(HOME_LOCAL, "03_physeq_LSP48_LSP49_genus_merged.rds", sep = "")) # Phyloseq object for 16S rRNA gene, merged on genus level
physeq2 <- readRDS(paste(HOME_LOCAL, "03_physeq_LSP08_LSP10_genus_merged.rds", sep = "")) # Phyloseq object for ITS region, merged on genus level
name <- "LSP4849_genus_merged_LSP0810_genus_merged"
level <- "Genus"
```

### Create meta data table

Input files are ITS UNITE taxonomy annotation and 16S SILVA taxonomy annotation.

```{r eval=FALSE}
library(dplyr)

## create meta data table

# merge samples, i.e. aggregate replicates and runs
physeq1_m <- merge_samples(x = physeq1, group = "Sample", fun = sum)
physeq2_m <- merge_samples(x = physeq2, group = "Sample", fun = sum)

# keep samples present in both objects
common <- intersect(sample_names(physeq1_m), sample_names(physeq2_m))
physeq1_m <- prune_samples(common, physeq1_m)
physeq2_m <- prune_samples(common, physeq2_m)

# add ITS annotation
metaITS <- data.frame(ID = 1:ntaxa(physeq2_m),
                      Name = data.frame(tax_table(physeq2_m))[,1],
                      Organism = "ITS")

Annot_ITS <- read.table(paste(HOME_LOCAL, "02_LSP08_LSP10_sub_bed_unite_species_assignment_dynamic_29.11.2022.txt", sep = ""),
                        header = TRUE, stringsAsFactors = F, sep = "\t", row.names = NULL)
Annot_ITS <- Annot_ITS %>% select(3:which(names(Annot_ITS) == level)) %>% distinct()
Annot_ITS <- Annot_ITS[!is.na(Annot_ITS[,level]),]
metaITS <- left_join(metaITS, Annot_ITS, by = c("Name" = level))

# add 16S annotation
meta16S <- data.frame(ID = (ntaxa(physeq2_m) + 1) : (ntaxa(physeq2_m) + ntaxa(physeq1_m)),
                      Name = data.frame(tax_table(physeq1_m))[,1],
                      Organism = "16S")

Annot_16S <- read.table(paste(HOME_LOCAL, "02_LSP48_LSP49_silva_species_assignment_v138.txt", sep = ""),
                        header = TRUE, stringsAsFactors = F, sep = "\t", row.names = NULL)
Annot_16S <- Annot_16S %>% select(4:which(names(Annot_16S) == level)) %>% distinct()
Annot_16S <- Annot_16S[!is.na(Annot_16S[,level]),]
meta16S <- left_join(meta16S, Annot_16S, by = c("Name" = level))

# merge annotations; taxa without genus assignment are labelled by kingdom
meta <- bind_rows(metaITS, meta16S)
meta$Name[is.na(meta$Name) & meta$Organism == "ITS"] <- "unassigned_fungal_genus"
meta$Name[is.na(meta$Name) & meta$Organism == "16S"] <- "unassigned_bacterial_genus"
write.table(meta, paste(HOME_LOCAL, "SPIEC-EASI_robustness/Meta_info_", name, ".txt", sep = ""), row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)
```

### Run network analysis

```{r eval=FALSE}
library(SpiecEasi)

set.seed(1)

## calculate networks for each rank and compute EBIC

ranks <- c(1, 2, 3, 6, 10, 11, 18, 32, 40, 45, 48, 50, 55)
ebic_table <- data.frame(rank = ranks, ebic = NA, degenerate = NA, opt_index = NA, edges = NA)

for(i in 1:length(ranks)){

  r <- ranks[i]

  # calculate slr
  phy.slr <- spiec.easi(list(physeq2_m, physeq1_m), method='slr', nlambda=50,
                        lambda.min.ratio=1e-2, r=r, lambda.log=TRUE,
                        pulsar.params=list(ncores=32, rep.num=30))   # as for the published models
  saveRDS(phy.slr, file = paste(HOME_LOCAL, "SPIEC-EASI_robustness/SLR_model_", name, "_rank", r, ".rds", sep = ""))

  # EBIC (log-determinant version, see 00_network_functions.R); degenerate fits are flagged and excluded
  ebic_table[i, "ebic"] <- ebic_logdet(phy.slr)
  ebic_table[i, "degenerate"] <- is_degenerate(phy.slr)
  ebic_table[i, "opt_index"] <- phy.slr$select$stars$opt.index
  ebic_table[i, "edges"] <- sum(getRefit(phy.slr)) / 2
}

write.table(ebic_table, file = paste(HOME_LOCAL, "SPIEC-EASI_robustness/SLR_model_", name, "_ebictable.txt", sep = ""),
            col.names = TRUE, row.names = FALSE, sep = "\t", quote = FALSE)

# ranks ordered by EBIC (non-degenerate fits)
ebic_table[!ebic_table$degenerate, ][order(ebic_table$ebic[!ebic_table$degenerate]), ]
```

# Robustness analysis

Networks estimated at the lambda selected for the full data, for five random subsamples of 80% of the samples.
Relative Hamming distance (FP + FN) / P (Kurtz et al. 2019) over all taxa of the full-data network, where FP = edges only in the subsample network, FN = edges of the full-data network not recovered (including edges of taxa without reads in the subsample), P = edges of the full-data network.

```{r eval=FALSE}
counts <- counts_from_physeq(physeq2, physeq1)   # ITS first, 16S second, as in spiec.easi(list(physeq2_m, physeq1_m))

# test ranks in the order of their EBIC until the first robust model is found
candidates <- ebic_table[!ebic_table$degenerate, ]
candidates <- candidates[order(candidates$ebic), ]
hamming_table <- data.frame()

for(i in 1:nrow(candidates)){
  r <- candidates$rank[i]
  nets <- fixed_lambda_networks(counts, r = r, opt = candidates$opt_index[i])
  h <- hamming_distances(nets)
  hamming_table <- rbind(hamming_table, data.frame(rank = r, ebic = candidates$ebic[i], subsample = names(h), hamming_dist = h))
  print(paste("rank", r, "mean relative Hamming distance", round(mean(h), 2)))
  if(mean(h) < 0.5) break
}

write.table(hamming_table, file = paste(HOME_LOCAL, "SPIEC-EASI_robustness/Hamming_dist_table_", name, ".txt", sep = ""),
            col.names = TRUE, row.names = FALSE, sep = "\t", quote = FALSE)
r_selected <- r   # rank 1 for this network (rank 40, the EBIC-optimal model, has a mean distance of 0.81)
```

# Export edge weights of the selected network

```{r eval=FALSE}
phy.slr <- readRDS(paste(HOME_LOCAL, "SPIEC-EASI_robustness/SLR_model_", name, "_rank", r_selected, ".rds", sep = ""))
edges <- selected_edges(phy.slr, meta$Name)
write.table(edges, paste(HOME_LOCAL, "SPIEC-EASI_robustness/Edge_weights_", name, "_rank", r_selected, ".txt", sep = ""),
            row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)

# summary: 165 edges, 134 negative
table(edges$sign)
```
