---
title: "Bacterial-fungal network calculation"
output:
  html_document:
    theme: united
    highlight: breezedark
    toc: yes
    toc_float: true
    df_print: paged
---

# Outline

- Calculate reference network with all taxa on an array of r (rank L), evaluate r that gives lowest ebic.

- Subsample taxa (80% per kingdom), calculate network with previously defined r.

- Calculate relative Hamming distance to reference network ( (False positive + False negative)/ true edges ).


# Calculate reference network

Calculate reference network with all taxa on an array of r (rank L).
Determine r that gives lowest ebic.

### Specify input

```{r eval=FALSE}
library(phyloseq)

## specify input files
physeq1 <- readRDS(paste(HOME_LOCAL, "physeq_LSP48_LSP49_genus_merged.rds", sep = "")) # Phyloseq object for 16S, merged on genus level
physeq2 <- readRDS(paste(HOME_LOCAL, "physeq_LSP08_LSP10_genus_merged.rds", sep = "")) # Phyloseq object for ITS, merged on genus level
name <- "LSP4849_genus_merged_LSP0810_genus_merged"
level <- "Genus"
```

### Create meta data table

Input files are ITS UNITE taxonomy annotation and 16S SILVA taxonomy annotation.

```{r eval=FALSE}
library(dplyr)

## create meta data table

# merge samples, i.e. aggregate replicates and runs
physeq1_m <- merge_samples(x = physeq1, group = "Sample", fun = sum)
physeq2_m <- merge_samples(x = physeq2, group = "Sample", fun = sum)


# add ITS annotation
metaITS <- data.frame(ID = 1:nrow(data.frame(tax_table(physeq2_m))), 
                       Name = data.frame(tax_table(physeq2_m))[,1],
                       Organism = "ITS")

Annot_ITS <- read.table(paste(HOME_LOCAL,"02_LSP08_LSP10_sub_bed_unite_species_assignment_dynamic_29.11.2022.txt", 
                             sep = "/"), header = TRUE, stringsAsFactors = F, sep = "\t", row.names = NULL)

Annot_ITS <- Annot_ITS %>% select(3:which(names(Annot_ITS) == level)) %>% distinct()
Annot_ITS <- Annot_ITS[!is.na(Annot_ITS[,level]),]

metaITS <- left_join(metaITS, Annot_ITS, by=c("Name" = level))


# add 16S annotation
meta16S <- data.frame(ID = (nrow(data.frame(tax_table(physeq2_m))) + 1) : (nrow(data.frame(tax_table(physeq2_m))) + nrow(data.frame(tax_table(physeq1_m)))), 
                       Name = data.frame(tax_table(physeq1_m))[,1],
                       Organism = "16S")

Annot_16S <- read.table(paste(HOME_LOCAL,"02_LSP48_LSP49_silva_species_assignment_v138.txt", sep = "/"), 
                       header = TRUE, stringsAsFactors = F, sep = "\t", row.names = NULL)
Annot_16S <- Annot_16S %>% select(4:which(names(Annot_16S) == level)) %>% distinct()
Annot_16S <- Annot_16S[!is.na(Annot_16S[,level]),]

meta16S <- left_join(meta16S, Annot_16S, by=c("Name" = level))

# merge annotations
meta <- rbind(metaITS, meta16S)
write.table(meta, paste(HOME_LOCAL,"SPIEC-EASI_robustness/Meta_info_", name, ".txt", sep = ""), row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)
```

### Run network analysis

```{r eval=FALSE}
library(SpiecEasi)
library(reshape2)

set.seed(1)

## calculate networks

# initialize summary table for ebic and rank

ranks <- c(1, 2,  3,  6, 10, 11, 18, 32, 40, 45, 48, 50, 55)

ebic_table <- data.frame(matrix(nrow = length(ranks), ncol = 2))
colnames(ebic_table) <- c("rank", "ebic")

# loop through ranks
for(i in 1:length(ranks)){
  
  r <- ranks[i]
  
  # calculate slr
  phy.slr <- spiec.easi(list(physeq2_m, physeq1_m), method='slr', nlambda=50,
           lambda.min.ratio=1e-2, r=r, lambda.log=TRUE,
           pulsar.params=list(ncores=32, rep.num=30))
  saveRDS(phy.slr, file = paste(HOME_LOCAL,"SPIEC-EASI_robustness/SLR_model_", name, "_rank", r, ".rds", sep = ""))
  
  # plot
  dtype <- c(rep(1,ntaxa(physeq2_m)), rep(2,ntaxa(physeq1_m)))
  pdf(file = paste(HOME_LOCAL,"SPIEC-EASI_robustness/SLR_model_", name, "_rank", r, ".pdf", sep = ""), width = 10, height = 10 )
  plot(adj2igraph(getRefit(phy.slr)), vertex.color=dtype+1, vertex.size=9)
  dev.off() 

  # get ebic
  eBic <- ebic(phy.slr$refit$stars, phy.slr$est$data, phy.slr$est$loglik[phy.slr$select$stars$opt.index])
  
  # save ranks a nd ebic in table
  ebic_table[i,"rank"] <- r
  ebic_table[i,"ebic"] <- eBic
  
  # obtain matrix with edge weights

  se.icov <- phy.slr$select$est$icov[[getOptInd(phy.slr)]]
  
  tryCatch({
  # Try to solve the matrix
  se.cov <<- solve(se.icov)
  # Continue with the rest of your code if solve is successful
}, error = function(e) {
  # if the matrix to solve is singular and does not have an inverse, add a small constant to thediagonal elements of the matrix. 
  # This is known as regularization or ridge regression in the context of linear regression.
   lambda <- 1e-6  # Small constant
   se.icov.regularized <- se.icov + diag(lambda, nrow(se.icov))
   se.cov <<- solve(se.icov.regularized)
})

  # format to long format
  df_long <- melt(se.cov, varnames = c("Source", "Target"), value.name = "Weight")

  df_long2 <- df_long %>%
  mutate(temp = pmin(as.numeric(Source), as.numeric(Target)),
         Target = pmax(as.numeric(Source), as.numeric(Target)),
         Source = temp) %>% 
  select(-temp) 

  df_long3 <- df_long2 %>% filter(!duplicated(df_long2[c("Source", "Target")])) %>%
  filter(Source != Target)
  write.table(df_long3, paste(HOME_LOCAL,"SPIEC-EASI_robustness/Edge_weigths_", name, "_rank", r, "_all.txt", sep = ""), 
              row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)

  # annotate long format with meta data table
  df_long3_a <- left_join(df_long3, meta[,c("ID", "Name")], by=c("Source" = "ID"))
  df_long3_a <- left_join(df_long3_a, meta[,c("ID", "Name")], by=c("Target" = "ID"))
  colnames(df_long3_a)[4:5] <- c("Source_name", "Target_name")
  write.table(df_long3_a, paste(HOME_LOCAL,"SPIEC-EASI_robustness/Edge_weigths_", name, "_rank", r, "_all_annot.txt", sep = ""), 
              row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)

}

# save ebic table
write.table(ebic_table, file = paste(HOME_LOCAL, "SPIEC-EASI_robustness/SLR_model_", name, "_ebictable.txt", sep = ""), 
            col.names = TRUE, row.names = FALSE, sep = "\t", quote = FALSE) 

# rank for min ebic
ebic_table[ebic_table$ebic==min(ebic_table$ebic),"rank" ]


```




# Robustness analysis

Calculate network on taxa subset and calculate Hamming distance to ref network.

Subsample taxa (80% per kingdom), calculate network with previously defined r.


```{r eval=FALSE}
library(phyloseq)
library(SpiecEasi)
library(dplyr)

## specify input 

# set genus
physeq1 <- readRDS(paste(HOME_LOCAL, "physeq_LSP48_LSP49_genus_merged.rds", sep = "")) # input physeq1
physeq2 <- readRDS(paste(HOME_LOCAL, "physeq_LSP08_LSP10_genus_merged.rds", sep = "")) # input physeq2
name <- "LSP4849_genus_merged_LSP0810_genus_merged" # name
level <- "Genus" # tax glom level
r <- 1 # specify optimal rank
ref_weights <- read.table(paste(HOME_LOCAL, "SPIEC-EASI_robustness/Edge_weigths_", name, "_rank", r, "_all_annot.txt" , sep = ""), 
                        header = TRUE, stringsAsFactors = FALSE) # input reference edges
n <- 50 # number of subnetworks to calculate



## sample and calculate i subnetworks, calculate hamming distance to ref

# initialize hamming output matrix
hamming_table <- data.frame(matrix(nrow = n, ncol = 2))
colnames(hamming_table) <- c("subnetwork_number", "hamming_dist")


for(i in 1:n){
  
  set.seed(i)
  
  ## subsample taxa

  # subsample physeq
  taxa_to_keep <- sample(taxa_names(physeq1), size = round(ntaxa(physeq1) * 0.8))
  physeq1_subset <- prune_taxa(taxa_to_keep, physeq1)
  
  taxa_to_keep <- sample(taxa_names(physeq2), size = round(ntaxa(physeq2) * 0.8))
  physeq2_subset <- prune_taxa(taxa_to_keep, physeq2)
  
  # merge samples, i.e. aggregate replicates and runs
  physeq1_m <- merge_samples(x = physeq1_subset, group = "Sample", fun = sum)
  physeq2_m <- merge_samples(x = physeq2_subset, group = "Sample", fun = sum)

  # check whether there are some samples missing in one or the other physeq
  # subset for samples present in both physeq objects
  sample_vec <- c(rownames(physeq1_m@sam_data), rownames(physeq2_m@sam_data))
  nondup <- sample_vec[!duplicated(sample_vec) & !duplicated(sample_vec, fromLast = TRUE)]
  
  if (length(nondup) > 0) {
     physeq1_m <- prune_samples( sample_names(physeq1_m) != nondup, physeq1_m)
     physeq2_m <- prune_samples( sample_names(physeq2_m) != nondup, physeq2_m)
  }
  

  ## calculate subnetwork with rank r
  
  # calculate slr
  phy.slr <- spiec.easi(list(physeq2_m, physeq1_m), method='slr', nlambda=50,
           lambda.min.ratio=1e-2, r=r, lambda.log=TRUE,
           pulsar.params=list(ncores=32, rep.num=30))
  saveRDS(phy.slr, file = paste(HOME_LOCAL,"SPIEC-EASI_robustness/SLR_model_", name, "_rank", r, "_subnetwork_", i, ".rds", sep = ""))
  
  # plot
  dtype <- c(rep(1,ntaxa(physeq2_m)), rep(2,ntaxa(physeq1_m)))
  pdf(file = paste(HOME_LOCAL,"SPIEC-EASI_robustness/SLR_model_", name, "_rank", r, "_subnetwork_", i, ".pdf", sep = ""), width = 10, height = 10 )
  plot(adj2igraph(getRefit(phy.slr)), vertex.color=dtype+1, vertex.size=9)
  dev.off() 

  
  ## get meta data table for subnetwork
  
  # ITS annotation
  metaITS <- data.frame(ID = 1:nrow(data.frame(tax_table(physeq2_m))), 
                       Name = data.frame(tax_table(physeq2_m))[,1],
                       Organism = "ITS")
  Annot_ITS <- read.table(paste(HOME_LOCAL,"02_LSP08_LSP10_sub_bed_unite_species_assignment_dynamic_29.11.2022.txt", 
                             sep = "/"), header = TRUE, stringsAsFactors = F, sep = "\t", row.names = NULL)
  Annot_ITS <- Annot_ITS %>% select(3:which(names(Annot_ITS) == level)) %>% distinct()
  Annot_ITS <- Annot_ITS[!is.na(Annot_ITS[,level]),]
  metaITS <- left_join(metaITS, Annot_ITS, by=c("Name" = level))

  # 16S annotation
  meta16S <- data.frame(ID = (nrow(data.frame(tax_table(physeq2_m))) + 1) : (nrow(data.frame(tax_table(physeq2_m))) + nrow(data.frame(tax_table(physeq1_m)))), 
                       Name = data.frame(tax_table(physeq1_m))[,1],
                       Organism = "16S")
  Annot_16S <- read.table(paste(HOME_LOCAL,"02_LSP48_LSP49_silva_species_assignment_v138.txt", sep = "/"), 
                       header = TRUE, stringsAsFactors = F, sep = "\t", row.names = NULL)
  Annot_16S <- Annot_16S %>% select(4:which(names(Annot_16S) == level)) %>% distinct()
  Annot_16S <- Annot_16S[!is.na(Annot_16S[,level]),]
  meta16S <- left_join(meta16S, Annot_16S, by=c("Name" = level))

  meta <- rbind(metaITS, meta16S)
  
  
  ## obtain matrix with edge weights

  se.icov <- phy.slr$select$est$icov[[getOptInd(phy.slr)]]
  
  tryCatch({
  # Try to solve the matrix
  se.cov <<- solve(se.icov)
  # Continue with the rest of your code if solve is successful
}, error = function(e) {
  # if the matrix to solve is singular and does not have an inverse, add a small constant to thediagonal elements of the matrix. 
  # This is known as regularization or ridge regression in the context of linear regression.
   lambda <- 1e-6  # Small constant
   se.icov.regularized <- se.icov + diag(lambda, nrow(se.icov))
   se.cov <<- solve(se.icov.regularized)
})

  # format to long format
  edges_long <- melt(se.cov, varnames = c("Source", "Target"), value.name = "Weight")

  edges_long <- edges_long %>%
  mutate(temp = pmin(as.numeric(Source), as.numeric(Target)),
         Target = pmax(as.numeric(Source), as.numeric(Target)),
         Source = temp) %>% 
  select(-temp) 

  edges_long <- edges_long %>% filter(!duplicated(edges_long[c("Source", "Target")])) %>%
  filter(Source != Target)
  write.table(edges_long, paste(HOME_LOCAL,"SPIEC-EASI_robustness/Edge_weigths_", name, "_rank", r, "_subnetwork_", i, "_all.txt", sep = ""), 
              row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)

  # annotate long format with meta data table
  edges_long_annot <- left_join(edges_long, meta[,c("ID", "Name")], by=c("Source" = "ID"))
  edges_long_annot <- left_join(edges_long_annot, meta[,c("ID", "Name")], by=c("Target" = "ID"))
  colnames(edges_long_annot)[4:5] <- c("Source_name", "Target_name")
  write.table(edges_long_annot, paste(HOME_LOCAL,"SPIEC-EASI_robustness/Edge_weigths_", name, "_rank", r, "_subnetwork_", i, "_all_annot.txt", sep = ""), 
              row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)

  
  ## calculate hamming distance
  
  # extract shared nodes
  ref_edges <- na.omit(ref_weights[,c(4:5,3)])
  ref_edges2 <- ref_edges
  ref_edges2[,1] <- pmin(ref_edges[,1], ref_edges[,2]); ref_edges2[,2] <- pmax(ref_edges[,1], ref_edges[,2]) # order alphabetically 
  
  sub_edges <- na.omit(edges_long_annot[,c(4:5,3)])
  sub_edges2 <- sub_edges
  sub_edges2[,1] <- pmin(sub_edges[,1], sub_edges[,2]); sub_edges2[,2] <- pmax(sub_edges[,1], sub_edges[,2]) # order alphabetically 
  
  edges <- left_join(sub_edges2, ref_edges2, by=c("Source_name" = "Source_name", "Target_name" = "Target_name"))
  colnames(edges)[3:4] <- c("Weight_sub", "Weight_ref")
  
  # calculate relative Hamming
  FP <- sum(edges$Weight_sub > 0 & edges$Weight_ref <= 0)
  FN <- sum(edges$Weight_sub < 0 & edges$Weight_ref >= 0)

  hamming_table[i,"subnetwork_number"] <- i
  hamming_table[i,"hamming_dist"] <- (FP + FN) / nrow(edges)
  
}

# save hamming summary table
write.table(hamming_table, file = paste(HOME_LOCAL, "SPIEC-EASI_robustness/Hamming_dist_table_", name, "_rank", r, ".txt", sep = ""), 
            col.names = TRUE, row.names = FALSE, sep = "\t", quote = FALSE) 

```

Visualize hamming distances

```{r eval=FALSE}
library(ggplot2)

## read hamming dist tables

file_names <- list.files(path = paste(HOME_LOCAL, "SPIEC-EASI_robustness/", sep = ""), 
                         pattern = "^Hamming_dist_table_", full.names = TRUE)

df_list <- lapply(file_names, function(file) {
  df <- read.table(file, header = TRUE)
  df$Set <- basename(file)
  return(df)
})

hamming_dist <- do.call(rbind, df_list)
hamming_dist$Set <- sub(x = hamming_dist$Set, pattern = "Hamming_dist_table_", replacement = "")
hamming_dist$Set <- sub(x = hamming_dist$Set, pattern = ".txt", replacement = "")

hamming_dist$Set <- factor(x = hamming_dist$Set, 
                               labels = c("Order", "Family", "Genus"))

## plot
p <- ggplot(hamming_dist, aes(x = Set, y = hamming_dist)) +
  geom_boxplot() +
  theme_minimal() +
  labs(x = "", y = "Relative hamming distance")
ggsave(paste(HOME_LOCAL, "SPIEC-EASI_robustness/Summary_plot_relative_hamming_dist_ITS_16S.pdf", sep = ""), plot = p, width = 3, height = 3)
```

