---
title: "Pseudomonas- Z.tritici network calculation"
output:
  html_document:
    theme: united
    highlight: breezedark
    toc: yes
    toc_float: true
    df_print: paged
---

# Outline

- Calculate reference network with all taxa on an array of r (rank L), evaluate r that gives lowest ebic.

- Subsample taxa (80% per kingdom), calculate network with previously defined r.

- Calculate relative Hamming distance to reference network ( (False positive + False negative)/ true edges ).

# Calculate reference network

Calculate reference network with all taxa on an array of r (rank L).
Determine r that gives lowest ebic.

### Specify input

```{r eval=FALSE}
library(phyloseq)

## specify input files
physeq1 <- readRDS(paste(HOME_LOCAL, "physeq_LSP36_LSP46_tipglom_0_top200.rds", sep = "")) # Phyloseq object for Pseudomonas, 200 most abundant ASV
physeq2 <- readRDS(paste(HOME_LOCAL, "physeq_LSP15_LSP16_tipglom_0_top200.rds", sep = "")) # Phyloseq object for Z. tritici, 200 most abundant ASV
name <- "LSP3646_0_top200_LSP1516_0_top200"
```

### Create meta data table

Input files are ITS UNITE taxonomy annotation and 16S SILVA taxonomy annotation.

```{r eval=FALSE}
library(dplyr)

### Create meta data table

# identify samples that are not contained in both physeq objects
sample_vec <- c(rownames(physeq1@sam_data), rownames(physeq2@sam_data))
nondup <- sample_vec[!duplicated(sample_vec) & !duplicated(sample_vec, fromLast = TRUE)]

# exclude samples that are not contained in both physeq objects
physeq1_m <- prune_samples( sample_names(physeq1) != nondup, physeq1)
physeq2_m <- prune_samples( sample_names(physeq2) != nondup, physeq2)


## make annotations

# PS annot
meta1 <- data.frame(ID = (length(colnames(otu_table(physeq2)))+1) : ( length(colnames(otu_table(physeq2))) + length(colnames(otu_table(physeq1)))), 
                    Name = colnames(physeq1_m@otu_table),
                    Species = tax_table(physeq1_m)[,"Species"],
                    Organism = "PS")
meta1$Name <- paste("P_", meta1$Name, sep = "")

# Zymo annot
meta2 <- data.frame(ID = 1:length(colnames(otu_table(physeq2))), Name = colnames(otu_table(physeq2)),
                    Species = colnames(otu_table(physeq2)),
                    Organism = "ZY", Group = "ZY", Subgroup = "ZY")
meta2$Name <- paste("Z_", meta2$Name, sep = "")

# add PS group and subgroup info from blast file  
annot <- read.table(paste(HOME_LOCAL, "LSP36_LSP46_blast_summary_2.txt", sep = ""), sep = "\t", header = TRUE, stringsAsFactors = FALSE)
annot <- annot[,c("Species", "Group", "Subgroup")] %>% distinct()
meta1 <- left_join(meta1, annot, by = c("Species" = "Species"))
meta1$Species <- sub(x = meta1$Species, pattern = "Pseudomonas", replacement = "P.")

# merge PS and Zymo annotations
meta <- rbind(meta2, meta1)
write.table(meta, paste(HOME_LOCAL,"SPIEC-EASI_robustness_PS_Zymo/Meta_info_", name, ".txt", sep = ""), row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)
print(paste("Note 2: The meta table of set", name, "finsihed."))

```

### Run network analysis

```{r eval=FALSE}
library(SpiecEasi)
library(reshape2)

set.seed(1)

# initialize summary table for ebic and rank

ranks <- c(1, 2,  3,  6, 10, 11, 18, 32, 40, 45, 48, 50, 55)

ebic_table <- data.frame(matrix(nrow = length(ranks), ncol = 2))
colnames(ebic_table) <- c("rank", "ebic")

# loop through ranks
for(i in 1:length(ranks)){
  
  r <- ranks[i]
  print(paste("Note 3: The calculation of ref network", name,  "with rank", r, "is now starting."))
  
  # calculate slr
  tryCatch({
  phy.slr <- spiec.easi(list(physeq2_m, physeq1_m), method='slr', nlambda=50,
                        lambda.min.ratio=1e-2, r=r, lambda.log=TRUE,
                        pulsar.params=list(ncores=4, rep.num=30))
  saveRDS(phy.slr, file = paste(HOME_LOCAL,"SPIEC-EASI_robustness_PS_Zymo/SLR_model_", name, "_rank", r, ".rds", sep = ""))
  
  # plot
  dtype <- c(rep(1,ntaxa(physeq2_m)), rep(2,ntaxa(physeq1_m)))
  pdf(file = paste(HOME_LOCAL,"SPIEC-EASI_robustness_PS_Zymo/SLR_model_", name, "_rank", r, ".pdf", sep = ""), 
      width = 10, height = 10 )
  plot(adj2igraph(getRefit(phy.slr)), vertex.color=dtype+1, vertex.size=9)
  dev.off() 
  
  # get ebic
  eBic <- ebic(phy.slr$refit$stars, phy.slr$est$data, phy.slr$est$loglik[phy.slr$select$stars$opt.index])
  
  }, error = function(e) {
    # if the network calculation fails, save eBic as NA
    eBic <- NA
    print(paste("Note 4.2: The calculation of ref network", name,  "with rank", r, "failed."))
  })
  
  # save ranks a nd ebic in table
  ebic_table[i,"rank"] <- r
  ebic_table[i,"ebic"] <- eBic
  
  if(!is.na(ebic)) {
  # obtain adjacency matrix
  
  # inverse covariance
  se.icov <- phy.slr$select$est$icov[[getOptInd(phy.slr)]]
  
  # covariance
  tryCatch({
    # try to solve the matrix
    se.cov <<- solve(se.icov)
  }, error = function(e) {
    # if the matrix to solve is singular and does not have an inverse, add a small constant to thediagonal elements of the matrix. 
    # This is known as regularization or ridge regression in the context of linear regression.
    lambda <- 1e-6  # Small constant
    se.icov.regularized <- se.icov + diag(lambda, nrow(se.icov))
    se.cov <<- solve(se.icov.regularized)
  })
  
  # calculate adjacency matrix 
  # specifically, refit precision matrix (inverse cov) to the data, multiply by covariance to get weighted covariance matrix
  adj <- as.matrix(se.cov*getRefit(phy.slr))
  
  # format to long format
  df_long <- melt(adj, varnames = c("Source", "Target"), value.name = "Weight")
  
  df_long2 <- df_long %>%
    mutate(temp = pmin(as.numeric(Source), as.numeric(Target)),
           Target = pmax(as.numeric(Source), as.numeric(Target)),
           Source = temp) %>% select(-temp) 
  
  df_long3 <- df_long2 %>% filter(!duplicated(df_long2[c("Source", "Target")])) %>% filter(Source != Target)
  write.table(df_long3, paste(HOME_LOCAL,"SPIEC-EASI_robustness_PS_Zymo/Edge_weigths_", name, "_rank", r, "_all.txt", sep = ""), 
              row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)
  
  # annotate long format with meta data table
  df_long3_a <- left_join(df_long3, meta[,c("ID", "Name")], by=c("Source" = "ID"))
  df_long3_a <- left_join(df_long3_a, meta[,c("ID", "Name")], by=c("Target" = "ID"))
  colnames(df_long3_a)[4:5] <- c("Source_name", "Target_name")
  write.table(df_long3_a, paste(HOME_LOCAL,"SPIEC-EASI_robustness_PS_Zymo/Edge_weigths_", name, "_rank", r, "_all_annot.txt", sep = ""), 
              row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)
  
  print(paste("Note 4.1: The calculation of ref network", name,  "with rank", r, "ran succesfully through."))
}}

# save ebic table
write.table(ebic_table, file = paste(HOME_LOCAL, "SPIEC-EASI_robustness_PS_Zymo/SLR_model_", name, "_ebictable.txt", sep = ""), 
            col.names = TRUE, row.names = FALSE, sep = "\t", quote = FALSE) 

# rank for min ebic
non_na_indices <- which(!is.na(ebic_table$ebic))
min_row <- non_na_indices[which.min(ebic_table$ebic[non_na_indices])]
r <- ebic_table$rank[min_row]

print(paste("Note 5: The optimization of ref network", name, "finsihed. The ebic table was saved. The optimal rank is", r))

```


# Robustness analysis

Calculate network on taxa subset and calculate Hamming distance to ref network.

Subsample taxa (80% per kingdom), calculate network with previously defined r.

```{r eval=FALSE}
 ## sample and calculate i subnetworks, calculate hamming distance to ref
    
   # initialize hamming output matrix
hamming_table <- data.frame(matrix(nrow = n, ncol = 2))
colnames(hamming_table) <- c("subnetwork_number", "hamming_dist")

# read in ref weights
ref_weights <- read.table(paste(HOME_LOCAL, "SPIEC-EASI_robustness_PS_Zymo/Edge_weigths_", name, 
                                "_rank", r, "_all_annot.txt" , sep = ""), 
                          header = TRUE, stringsAsFactors = FALSE) # input reference edges

for(i in 1:n){
  
  set.seed(i)
  
  ## subsample otu table based on samples
  
  samples_to_keep <- sample(sample_names(physeq1_m), size = round(nsamples(physeq1_m) * 0.8))
  
  print(paste("Note 6: Start the subsampling of refnetwork", name, "with rank", r, " - This is subsampling round", i))
  print(paste("Physeq1 has total",  nsamples(physeq1_m), "samples. Subsample", round(nsamples(physeq1_m) * 0.8), "samples"))
  
  physeq1_subset <- prune_samples(samples_to_keep, physeq1_m)
  physeq2_subset <- prune_samples(samples_to_keep, physeq2_m)
  
  ## calculate subnetwork with rank r
  
  # calculate slr
  tryCatch({
  phy.slr <- spiec.easi(list(physeq2_subset, physeq1_subset), method='slr', nlambda=50,
                        lambda.min.ratio=1e-2, r=r, lambda.log=TRUE,
                        pulsar.params=list(ncores=4, rep.num=30))
  saveRDS(phy.slr, file = paste(HOME_LOCAL,"SPIEC-EASI_robustness_PS_Zymo/SLR_model_", name, 
                                "_rank", r, "_subnetwork_", i, ".rds", sep = ""))
  
  # plot
  dtype <- c(rep(1,ntaxa(physeq2_subset)), rep(2,ntaxa(physeq1_subset)))
  pdf(file = paste(HOME_LOCAL,"SPIEC-EASI_robustness_PS_Zymo/SLR_model_", name, 
                   "_rank", r, "_subnetwork_", i, ".pdf", sep = ""), width = 10, height = 10 )
  plot(adj2igraph(getRefit(phy.slr)), vertex.color=dtype+1, vertex.size=9)
  dev.off() 
  
  
  ## get meta data table for subnetwork
  
  # PS annot
  meta1 <- data.frame(ID = (length(colnames(otu_table(physeq2_subset)))+1) : 
                        ( length(colnames(otu_table(physeq2_subset))) + length(colnames(otu_table(physeq1_subset)))), 
                      Name = colnames(physeq1_subset@otu_table),
                      Species = tax_table(physeq1_subset)[,"Species"],
                      Organism = "PS")
  meta1$Name <- paste("P_", meta1$Name, sep = "")
  
  # Zymo annot
  meta2 <- data.frame(ID = 1:length(colnames(otu_table(physeq2_subset))), Name = colnames(otu_table(physeq2_subset)),
                      Species = colnames(otu_table(physeq2_subset)),
                      Organism = "ZY", Group = "ZY", Subgroup = "ZY")
  meta2$Name <- paste("Z_", meta2$Name, sep = "")
  
  # add PS group and subgroup info from blast file  
  annot <- read.table(paste(HOME_LOCAL, "LSP36_LSP46_blast_summary_2.txt", sep = ""), sep = "\t", 
                      header = TRUE, stringsAsFactors = FALSE)
  annot <- annot[,c("Species", "Group", "Subgroup")] %>% distinct()
  meta1 <- left_join(meta1, annot, by = c("Species" = "Species"))
  meta1$Species <- sub(x = meta1$Species, pattern = "Pseudomonas", replacement = "P.")
  
  # merge PS and Zymo annotations
  meta <- rbind(meta2, meta1)
  
  
  # obtain adjacency matrix
  
  # inverse covariance
  se.icov <- phy.slr$select$est$icov[[getOptInd(phy.slr)]]
  
  # covariance
  tryCatch({
    # try to solve the matrix
    se.cov <<- solve(se.icov)
  }, error = function(e) {
    # if the matrix to solve is singular and does not have an inverse, add a small constant to thediagonal elements of the matrix. 
    # This is known as regularization or ridge regression in the context of linear regression.
    lambda <- 1e-6  # Small constant
    se.icov.regularized <- se.icov + diag(lambda, nrow(se.icov))
    se.cov <<- solve(se.icov.regularized)
  })
  
  # calculate adjacency matrix 
  # specifically, refit precision matrix (inverse cov) to the data, multiply by covariance to get weighted covariance matrix
  adj <- as.matrix(se.cov*getRefit(phy.slr))
  
  # format to long format
  edges_long <- melt(adj, varnames = c("Source", "Target"), value.name = "Weight")
  
  edges_long <- edges_long %>%
    mutate(temp = pmin(as.numeric(Source), as.numeric(Target)),
           Target = pmax(as.numeric(Source), as.numeric(Target)),
           Source = temp) %>% select(-temp) 
  
  edges_long <- edges_long %>% filter(!duplicated(edges_long[c("Source", "Target")])) %>% filter(Source != Target)
  write.table(edges_long, paste(HOME_LOCAL,"SPIEC-EASI_robustness_PS_Zymo/Edge_weigths_", name, 
                                "_rank", r, "_subnetwork_", i, "_all.txt", sep = ""), 
              row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)
  
  # annotate long format with meta data table
  edges_long_annot <- left_join(edges_long, meta[,c("ID", "Name")], by=c("Source" = "ID"))
  edges_long_annot <- left_join(edges_long_annot, meta[,c("ID", "Name")], by=c("Target" = "ID"))
  colnames(edges_long_annot)[4:5] <- c("Source_name", "Target_name")
  write.table(edges_long_annot, paste(HOME_LOCAL,"SPIEC-EASI_robustness_PS_Zymo/Edge_weigths_", name, 
                                      "_rank", r, "_subnetwork_", i, "_all_annot.txt", sep = ""), 
              row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)
  
  
  ## calculate hamming distance
  
  # extract shared nodes
  ref_edges <- na.omit(ref_weights[,c(4:5,3)])
  ref_edges2 <- ref_edges
  ref_edges2[,1] <- pmin(ref_edges[,1], ref_edges[,2])
  ref_edges2[,2] <- pmax(ref_edges[,1], ref_edges[,2]) # order alphabetically 
  
  sub_edges <- na.omit(edges_long_annot[,c(4:5,3)])
  sub_edges2 <- sub_edges
  sub_edges2[,1] <- pmin(sub_edges[,1], sub_edges[,2])
  sub_edges2[,2] <- pmax(sub_edges[,1], sub_edges[,2]) # order alphabetically 
  
  edges <- left_join(sub_edges2, ref_edges2, by=c("Source_name" = "Source_name", "Target_name" = "Target_name"))
  colnames(edges)[3:4] <- c("Weight_sub", "Weight_ref")
  
  # calculate relative Hamming
  FP <- sum(edges$Weight_sub > 0 & edges$Weight_ref <= 0)
  FN <- sum(edges$Weight_sub < 0 & edges$Weight_ref >= 0)
  
  # subset for non empty ref edges
  edges <- edges[edges$Weight_ref!=0,]
  
  hamming_table[i,"subnetwork_number"] <- i
  hamming_table[i,"hamming_dist"] <- (FP + FN) / nrow(edges)
  
  print(paste("Note 7.1: The subsampling of refnetwork", name, "with rank", r, "with subsampling round", i, "ran succesfully through."))
  
  }, error = function(e) {
    # if the network calculation fails, save Hamming as FAIL
    hamming_table[i,"subnetwork_number"] <- i
    hamming_table[i,"hamming_dist"] <- "FAIL"
    print(paste("Note 7.2: The subsampling of refnetwork", name, "with rank", r, "with subsampling round", i, "failed."))
  })
}

# save hamming summary table
write.table(hamming_table, file = paste(HOME_LOCAL, "SPIEC-EASI_robustness_PS_Zymo/Hamming_classic/Hamming_dist_table_", name, 
                                        "_rank", r, ".txt", sep = ""), 
            col.names = TRUE, row.names = FALSE, sep = "\t", quote = FALSE) 

print(paste("Note 8: The hamming table of refnetwork", name, "with rank", r, " was written."))
 
```

