---
title: "Pseudomonas-Z. tritici network calculation"
output:
  html_document:
    theme: united
    highlight: breezedark
    toc: yes
    toc_float: true
    df_print: paged
---

# Outline

- Calculate the network with all taxa for each SLR rank r (number of latent factors) and compute the EBIC.

- Export the StARS-selected edges with partial correlations as edge weights (Supplementary Table 13).

- Robustness: estimate networks on five random subsamples of 80% of the samples at the lambda selected for the full data, and calculate the relative Hamming distance (FP + FN) / P to the full-data network (Supplementary Table 14).

- The rank with the lowest EBIC among robust models (mean relative Hamming distance < 0.5) is used.

```{r eval=FALSE}
HOME_LOCAL <- "./"   # folder with the input files; output is written to HOME_LOCAL/SPIEC-EASI_robustness_PS_Zymo/
source(paste(HOME_LOCAL, "00_network_functions.R", sep = ""))
dir.create(paste(HOME_LOCAL, "SPIEC-EASI_robustness_PS_Zymo", sep = ""), showWarnings = FALSE)
```

# Calculate networks

### Specify input

```{r eval=FALSE}
library(phyloseq)

## specify input files
physeq1 <- readRDS(paste(HOME_LOCAL, "03_physeq_LSP36_LSP46_tipglom_0_top200.rds", sep = "")) # Phyloseq object for Pseudomonas, 200 most abundant ASVs (one sample per leaf)
physeq2 <- readRDS(paste(HOME_LOCAL, "03_physeq_LSP15_LSP16_tipglom_0_top200.rds", sep = "")) # Phyloseq object for Z. tritici, 200 most abundant ASVs (one sample per leaf)
name <- "LSP3646_0_top200_LSP1516_0_top200"
```

### Create meta data table

```{r eval=FALSE}
library(dplyr)

# keep samples present in both objects
common <- intersect(sample_names(physeq1), sample_names(physeq2))
physeq1_m <- prune_samples(common, physeq1)
physeq2_m <- prune_samples(common, physeq2)

# Zymo annotation
meta2 <- data.frame(ID = 1:ntaxa(physeq2_m), Name = paste("Z_", taxa_names(physeq2_m), sep = ""),
                    Species = taxa_names(physeq2_m), Organism = "ZY", Group = "ZY", Subgroup = "ZY")

# PS annotation, with group and subgroup from the blast file
meta1 <- data.frame(ID = (ntaxa(physeq2_m) + 1) : (ntaxa(physeq2_m) + ntaxa(physeq1_m)),
                    Name = paste("P_", taxa_names(physeq1_m), sep = ""),
                    Species = as.character(tax_table(physeq1_m)[,"Species"]),
                    Organism = "PS")
annot <- read.table(paste(HOME_LOCAL, "02_LSP36_LSP46_blast_summary_2.txt", sep = ""), sep = "\t", header = TRUE, stringsAsFactors = FALSE)
annot <- annot[,c("Species", "Group", "Subgroup")] %>% distinct()
meta1 <- left_join(meta1, annot, by = c("Species" = "Species"))
meta1$Species <- sub(x = meta1$Species, pattern = "Pseudomonas", replacement = "P.")

# merge PS and Zymo annotations
meta <- rbind(meta2, meta1)
write.table(meta, paste(HOME_LOCAL, "SPIEC-EASI_robustness_PS_Zymo/Meta_info_", name, ".txt", sep = ""), row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)
```

### Run network analysis

```{r eval=FALSE}
library(SpiecEasi)

set.seed(1)

## calculate networks for each rank and compute EBIC

ranks <- c(1, 2, 3, 6, 10, 11, 18, 32, 40, 45, 48, 50, 55)
ebic_table <- data.frame(rank = ranks, ebic = NA, degenerate = NA, opt_index = NA, edges = NA)

for(i in 1:length(ranks)){

  r <- ranks[i]

  # calculate slr
  phy.slr <- spiec.easi(list(physeq2_m, physeq1_m), method='slr', nlambda=50,
                        lambda.min.ratio=1e-2, r=r, lambda.log=TRUE,
                        pulsar.params=list(ncores=4, rep.num=30))   # as for the published models
  saveRDS(phy.slr, file = paste(HOME_LOCAL, "SPIEC-EASI_robustness_PS_Zymo/SLR_model_", name, "_rank", r, ".rds", sep = ""))

  # EBIC (log-determinant version, see 00_network_functions.R); degenerate fits are flagged and excluded
  ebic_table[i, "ebic"] <- ebic_logdet(phy.slr)
  ebic_table[i, "degenerate"] <- is_degenerate(phy.slr)
  ebic_table[i, "opt_index"] <- phy.slr$select$stars$opt.index
  ebic_table[i, "edges"] <- sum(getRefit(phy.slr)) / 2
}

write.table(ebic_table, file = paste(HOME_LOCAL, "SPIEC-EASI_robustness_PS_Zymo/SLR_model_", name, "_ebictable.txt", sep = ""),
            col.names = TRUE, row.names = FALSE, sep = "\t", quote = FALSE)

# ranks ordered by EBIC (non-degenerate fits)
ebic_table[!ebic_table$degenerate, ][order(ebic_table$ebic[!ebic_table$degenerate]), ]
```

# Robustness analysis

Networks estimated at the lambda selected for the full data, for five random subsamples of 80% of the samples.
Relative Hamming distance (FP + FN) / P (Kurtz et al. 2019) over all taxa of the full-data network, where FP = edges only in the subsample network, FN = edges of the full-data network not recovered (including edges of taxa without reads in the subsample), P = edges of the full-data network.

```{r eval=FALSE}
counts <- counts_from_physeq(physeq2, physeq1, merge_first = FALSE, merge_second = FALSE)   # Z. tritici first, Pseudomonas second

# test ranks in the order of their EBIC until the first robust model is found
candidates <- ebic_table[!ebic_table$degenerate, ]
candidates <- candidates[order(candidates$ebic), ]
hamming_table <- data.frame()

for(i in 1:nrow(candidates)){
  r <- candidates$rank[i]
  nets <- fixed_lambda_networks(counts, r = r, opt = candidates$opt_index[i])
  h <- hamming_distances(nets)
  hamming_table <- rbind(hamming_table, data.frame(rank = r, ebic = candidates$ebic[i], subsample = names(h), hamming_dist = h))
  print(paste("rank", r, "mean relative Hamming distance", round(mean(h), 2)))
  if(mean(h) < 0.5) break
}

write.table(hamming_table, file = paste(HOME_LOCAL, "SPIEC-EASI_robustness_PS_Zymo/Hamming_dist_table_", name, ".txt", sep = ""),
            col.names = TRUE, row.names = FALSE, sep = "\t", quote = FALSE)
r_selected <- r   # rank 45 for this network
# Note: for this network StARS instability stays below 0.05 along the whole lambda path, so the smallest lambda
# of the path (lambda.min.ratio = 1e-2) is selected (opt_index = 50).
```

# Export edge weights of the selected network

```{r eval=FALSE}
phy.slr <- readRDS(paste(HOME_LOCAL, "SPIEC-EASI_robustness_PS_Zymo/SLR_model_", name, "_rank", r_selected, ".rds", sep = ""))
edges <- selected_edges(phy.slr, meta$Name)
write.table(edges, paste(HOME_LOCAL, "SPIEC-EASI_robustness_PS_Zymo/Edge_weights_", name, "_rank", r_selected, ".txt", sep = ""),
            row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)

# summary: 917 edges, 29 negative
table(edges$sign)
```
